Experience Report: Haskell in Computational Biology

### 2012-07-01 mrfy_experience_report.tex
correcter 1-core data, rewrote perf section

\category{D.1.1}{Applicative (Functional) Programming}{}
\category{J.3}{Biology and genetics}{}
% \category{CR-number}{subcategory}{third-level}
% 
% \terms
% term1, term2
% 
% \keywords
% keyword1, keyword2

@@

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
Random hill climbing took 50~lines of code and one~day to
implement.  
Simulated annealing requires only a new
\texttt{utility} function, which took 15~lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm reuses the same
\texttt{pt0} and \texttt{SearchStop} functions, and its \texttt{utility}
function took only a few lines of code.
\remark{This claim isn't supported by inspection of the code.}
But~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.

@@

But \mrfy's job is not to run Viterbi's algorithm on large models;
\mrfy's job is to detect homologies to structures on which Viterbi's
algorithm is intractable.
We~benchmarked \mrfy\ 
using a model of an ``8-bladed beta propeller,''
which has 343 nodes, of which 178 appear in 40~beta strands.
The models between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
CAN WE PLEASE REPLACE THE TEXT ABOVE WITH A DESCRIPTION OF WHAT'S
GOING ON IN THE CODE WE ACTUALLY RAN???
On~these smaller problems, 
performance is good enough that \mrfy\ can do its job.
%%
%%
%%
SUDDEN CHANGE OF BENCHMARK CONSIDERED HARMFUL.
%%
%%
Given a Markov random field with a 12-stranded ``beta sandwich''
structure,
\mrfy\ typically computes an alignment within a minute;
SMURF allocates over 16GB of memory
and does not terminate even after eight hours.

Because \mrfy\ works on many subproblems independently, we~found it easy
to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
Using
\texttt{Control.Parallel}, parallelizing this computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
On a 48-core, 2.3GHz AMD Opteron 6176 system, 
parallel performance is good up to about 12~cores.
\figref{fig:speedup} shows speedup when
using a model of an ``8-bladed beta propeller'' with 1~to~48 cores.
The model has 165 ordinary nodes and 40~beta strands totalling
178 nodes; the benchmark query sequence has
592 amino acids.
AND WHAT DO WE CONCLUDE FROM THE FIGURE, JOHNNY?

@@

Anonymous referees spurred us to think
more deeply about laziness and to write more carefully about performance.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
Koen Claessen instantly diagnosed an inappropriately strict
implementation of the \texttt{Rand} monad.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci.

>>>>

\category{D.1.1}{Applicative (Functional) Programming}{}
\category{J.3}{Biology and genetics}{}
% \category{CR-number}{subcategory}{third-level}
% 
% \terms
% term1, term2
% 
\keywords
memoization, stochastic search, QuickCheck, homology detection

@@

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
Random hill climbing took 50~lines of code and one~day to
implement.  
Simulated annealing requires only a new
\texttt{utility} function, which took 15~lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm is very similar in three out of its four
functions,
but~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.

@@

But \mrfy's job is not to run Viterbi's algorithm on large models;
\mrfy's job is to detect homologies to structures on which Viterbi's
algorithm, and SMURF's more complex algorithm, are unsuited.
SMURF may be faster than \mrfy\ in some cases,
but \mrfy\ can solve problems that SMURF cannot.
\mrfy\ can compute an alignment to a complex, 12-stranded ``beta sandwich''
structure within one minute, while on the same structure, SMURF allocates
over 16GB of memory and does not terminate even after eight ours.
The Markov random field model for this structure contains 252 nodes,
97 of which appear in the 12 beta strands.
The models between beta strands typically have at most 12~nodes.
We used a query sequence of 237 amino acids, but each placement
breaks the sequence into 13 pieces, each of which typically has at most
11~amino acids.
On these smaller problems, performance is good enough that \mrfy\ can
do its job.

We benchmarked \mrfy\ using a model of an ``8-bladed beta propeller,''
which has 343 nodes, of which 178 appear in 40~beta strands.
The models between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a very high degree of parallelism.
And, in Haskell, we~found it easy
to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
Using
\texttt{Control.Parallel}, parallelizing this computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
On a 48-core, 2.3GHz AMD Opteron 6176 system, 
parallel performance is good up to about 12~cores.
\figref{fig:speedup} shows speedup when
using a model of an ``8-bladed beta propeller'' with 1~to~48 cores.
We can conclude that, at least on this particular system,
a single instance of \mrfy\ scales well to about 12~cores.
However, typical usage of \mrfy\ includes running many searches
in parallel, which gives us a way to use all 48 cores on this system.

@@

Anonymous referees spurred us to think
more deeply about laziness and to write more carefully about performance.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
Koen Claessen instantly diagnosed an inappropriately strict
implementation of the \texttt{Rand} monad.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci.
Noah Daniels and Andrew Gallant were funded in part by NIH grant 1R01GM080330 (to Lenore Cowen).
